---
title: "Lab 12 - Smoking during pregnancy"
author: "Sophie Boyd"
date: "4-10-26"
output: github_document
---

### Load packages and data

```{r load-packages, message=FALSE, warning=FALSE}
library(tidyverse) 
library(tidymodels)
library(openintro)
```

### Exercise 1

The numeric variables are fage, mage, weeks, visits, gained, and weight.

```{r numeric-summaries, warning=FALSE}
ncbirths %>%
  select(fage, mage, weeks, visits, gained, weight) %>%
  summarize(across(everything(), list(mean = mean, sd = sd), na.rm = TRUE))

ncbirths_long <- ncbirths %>%
  pivot_longer(
    cols = c("fage", "mage", "weeks", "visits", "gained", "weight"),
    names_to = "numeric_label",
    values_to = "numeric_value")

ncbirths_long %>%
  ggplot(aes(x = numeric_value)) +
  geom_boxplot() +
  facet_wrap(~numeric_label, nrow = 6) + 
  labs(x = NULL) 
```

Distributions are mostly balanced, with some high outliers on the amount of weight gained during pregnancy and low outliers on the number of weeks of pregnancy (which I assume are labeled as premature births with the "premie" variable).

### Exercise 2

```{r white-df}
ncbirths_white <- ncbirths %>%
  filter(whitemom %in% c("white"))

mean(ncbirths_white$weight)
```

### Exercise 3

- I believe that the observations in the dataset are independent of one another, assuming that each birth corresponds to a different mother. If multiple births from a mother with multiple children were featured in the dataset, this would violate the assumption of independence. I would need more information to verify.

- ncbirths_white has 714 observations, which I believe would be a reasonably large sample size for bootstrapping.

- In the boxplots from Exercise 1, I do not see any extreme skew or clustering that would cause problems. 


