---
title: "Lab 12 - Smoking during pregnancy"
author: "Sophie Boyd"
date: "4-10-26"
output: github_document
---

### Load packages and data

```{r load-packages, message=FALSE, warning=FALSE}
library(tidyverse) 
library(tidymodels)
library(openintro)
```

### Exercise 1

The numeric variables are fage, mage, weeks, visits, gained, and weight.

```{r numeric-summaries, warning=FALSE}
ncbirths %>%
  select(fage, mage, weeks, visits, gained, weight) %>%
  summarize(across(everything(), list(mean = mean, sd = sd), na.rm = TRUE))

ncbirths_long <- ncbirths %>%
  pivot_longer(
    cols = c("fage", "mage", "weeks", "visits", "gained", "weight"),
    names_to = "numeric_label",
    values_to = "numeric_value")

ncbirths_long %>%
  ggplot(aes(x = numeric_value)) +
  geom_boxplot() +
  facet_wrap(~numeric_label, nrow = 6) + 
  labs(x = NULL) 
```

Distributions are mostly balanced, but there is at least one outlier on each variable. The collections of high outliers on weight gained during pregnancy and low outliers on weeks and birth weight might be problematic for future analyses.


### Exercise 2

```{r white-df}
ncbirths_white <- ncbirths%>%
  filter(whitemom %in% c("white"))

mean(ncbirths_white$weight)
```

### Exercise 3

- I believe that the observations in the dataset are independent of one another, assuming that each birth corresponds to a different mother. If multiple births from a mother with multiple children were featured in the dataset, this would violate the assumption of independence. I would need more information to verify.

- ncbirths_white has 714 observations, which I believe would be a reasonably large sample size for bootstrapping.

- In the box plot from Exercise 1, I see a collection of low outliers, as well as a couple of high outliers, that might influence the results when analyzing birth weight. I chose to restrict my range to 3 standard deviations above and below the mean:

```{r birth-weight-chart}
ncbirths %>%
  ggplot(aes(x = weight)) +
  geom_boxplot() +
  labs(title = 'Distribution of birth weights before filtering to deal with outliers')

ncbirths_filtered <- ncbirths %>%
  filter(weight <= 11.63) %>%
  filter(weight >= 2.57)

ncbirths_filtered %>%
  ggplot(aes(x = weight)) +
  geom_boxplot() +
  labs(title = 'Distribution of birth weights after filtering to deal with outliers')
```

Now I need to update my dataframe before running the bootstrapping simulation:

```{r white-df-filtered}
ncbirths_white <- ncbirths_filtered%>%
  filter(whitemom %in% c("white"))

mean(ncbirths_white$weight)
```

The mean is slightly higher after removing the outliers. The sample size decreased from 714 to 703, so not a substantial change. 

### Exercise 4a

```{r bootstrapping-sim}
set.seed(123)

boot_df <- ncbirths_white %>%
  specify(response = weight) %>%
  generate(reps = 1000, type = "bootstrap") %>%
  calculate(stat = "mean")
```

### Exercise 4b 

```{r recenter}
mean(boot_df$stat)

boot_df <- boot_df %>%
  mutate(stat_cent1 = stat - 7.32) %>%
  mutate(stat_cent = stat_cent1 + 7.43)
```

### Exercise 4c

```{r sim-histogram, message=FALSE}
boot_df %>%
  ggplot(aes(x = stat_cent)) +
  geom_histogram() +
  geom_vline(aes(xintercept = 7.32),
             color = "blue",
             linetype = "dashed",
             linewidth = 1)
```

### Exercise 4d

```{r sig-test}
boot_df %>%
  summarize(lower = quantile(stat_cent, 0.025),
            upper = quantile(stat_cent, 0.975))
```

The mean of the simulated means fell just outside of the 95% confidence interval, meaning that just about 5% of the simulated means were at least as extreme as my observed value.

### Exercise 4e

The mean birth weight was significantly lower than 7.43 pounds, informing the conclusion that average birth weight for white babies has decreased significantly since 1995. 

### Exercise 5

```{r habit-weight-boxplot}
ncbirths_filtered %>%
  ggplot(aes(x = weight)) +
  geom_boxplot() +
  facet_wrap(~habit, nrow = 3)
```

Distributions of birth weights for babies of smoking and non-smoking mothers were similar, but the average birth weight was higher for babies of non-smokers than babies of smokers. Both distributions were somewhat skewed to the left. 

### Exercise 6

```{r ncbirths-clean}
ncbirths_clean <- ncbirths_filtered %>%
  filter(!is.na(habit)) %>%
  filter(!is.na(weight))
```

It is important to clean the data before obtaining group summaries because participants with missing data on "habit" will not fall under either group of interest and participants with missing data on "weight" do not provide information about the main dependent variable of interest.

### Exercise 7

```{r weight-diff}
ncbirths_clean %>%
  group_by(habit) %>%
  summarize(mean_weight = mean(weight))
```

7.26 - 6.91 = 0.35. The average difference in birth weight for non-smoking vs. smoking mothers is 0.35 pounds.

### Exercise 8

- Null hypothesis: The difference in the average birth weights of babies of non-smoking vs. smoking mothers is not significantly different from 0. (H0: µ1 - µ2 = 0)

- Alternative hypothesis: The difference in the average birth weights of babies of non-smoking vs. smoking mothers is significantly different from 0. (H1: µ1 - µ2 =/= 0)

### Exercise 9

To test whether maternal smoking is associated with a difference in birth weight, we could simulate a distribution of mean differences in birth weight between babies of non-smoking and smoking mothers to see if the simulated mean was significantly different from 0. 

- I think permutation would be the most appropriate method for this kind of hypothesis testing. **I need to review the ODD in the module and come back to this!

- The observed test statistic is the sample mean difference in birth weight between babies of non-smoking vs. smoking mothers (0.35).

- 
